% ============================================================
\section*{Generative AI Usage Statement}
% ============================================================

The author utilized Large Language Models for \LaTeX{} formatting, grammar correction, and copy-editing. The Constraint Cascade taxonomy, the four-layer architecture, the friction seam analysis, the cross-layer failure pattern taxonomy (Override, Inadequate Process Model, Compression), and all case analyses are original contributions of the author. No LLM was used to generate scientific claims or novel theoretical arguments.

% ============================================================
\section*{Ethical Considerations Statement}
% ============================================================

This paper discusses four real-world cases involving loss of life and serious harm. \emph{Garcia v.\ Character Technologies} involves a minor's suicide; \emph{Soelberg v.\ OpenAI} involves allegations arising from a murder-suicide by a person of diminished capacity; the Claude-Maven case involves reported military targeting operations. To respect the privacy and dignity of victims, all case details are drawn exclusively from public court filings, public company statements, and published reporting. No private investigation or unauthorized data collection was conducted.

The Tencent Yuanbao analysis draws exclusively from public social media posts, official corporate statements, and published news reports. Translations from Mandarin Chinese by the author; original text preserved where English equivalents are contested.
